Efficient pedigree recording for fast population genetics simulation

### 2018-06-07 forwards_paper.tex
ashander edits

\begin{abstract}
    In this paper we describe how to efficiently record the entire genetic history of a population
    in forwards-time, individual-based population genetics simulations with arbitrary breeding models,
    population structure and demography.
    This approach dramatically reduces the computational burden of tracking individual genomes
    by allowing us to simulate only those loci that may affect reproduction
    (those having non-neutral variants).
    The genetic history of the population is recorded as a succinct tree sequence
    as introduced in the software package \msprime,
    on which neutral mutations can be quickly placed afterwards.
    Recording the results of each breeding event
    results in a linear growth in the amount of information that must be stored.
    However, there is a great deal of redundancy in this information,
    since the majority of individuals simulated over the course of the simulation
    will leave no descendents in the eventual sample.
    We solve this storage problem by providing an algorithm to quickly `simplify'
    a tree sequence by removing this irrelevant history for a given set of genomes.
    By periodically simplifying the history with respect to the extant population,
    we show that the total storage space required is modest
    and overall large efficiency gains can be made over classical forward-time simulations.
    We implement a general-purpose framework for recording and simplifying genealogical data,
    which can be used to make simulations of any population model more efficient.
    We modify two popular forwards-time simulation frameworks to use this new approach
    and observe efficiency gains in large, whole-genome simulations of one to two orders of magnitude.
    In addition to speed, our method for recording pedigrees has several advantages:
    (1) All marginal genealogies of the simulated individuals are recorded, rather than just genotypes.
    (2) A population of $N$ individuals with $M$ polymorphic sites
        can be stored in $O(N \log N + M)$ space,
        making it feasible to store a simulation's entire final generation
        (and its history) rather than only a subset.
    (3) A simulation can easily be initialized with a more efficient coalescent simulation 
        of deep history.
    The software for recording and processing tree sequences is named \tskit.
\end{abstract}

@@

\paragraph{Essential steps in tree sequence recording}
Since a tree sequence can record the history of genetic inheritance in any situation
(requiring only unambiguous inheritance and no time travel),
any individual-based population genetics simulator can maintain a tree sequence
with only a little bookkeeping. \revpoint{1}{1}
We have furthermore provided several tools to minimize this bookkeeping,
only requiring a little one-way output at the birth of each new individual.
Concretely, to record a tree sequence, including mutations,
a simulator must record for each new genome
(so, twice for each new diploid individual): \revpoint{1}{6}
\begin{itemize}
    \item the birth time of the genome in the Node Table,
    \item the segments the genome inherits from its parental genomes in the Edge Table,
    \item the locations of any new mutations in the Site Table,
    \item and the derived state of these new mutations in the Mutation Table
        (as well as the identiy of this genome the mutations appeared in).
\end{itemize}
Each of these can be simply appended to the ends of the respective tables.
Besides this, simplification should be run every once in a while (e.g., every 100 generations).
Before simplification, time in the Node Table must be translated to ``time ago'' if it is not already.
(This was avoided in Algorithm~\algref{W} since there $t$ denoted
``time until the end of the simulation''.)
There are also several ``cleaning'' steps, for which we provide functions in the Tables API:
\emph{sorting} according to several criteria for algorithmic efficiency;
and removing any duplicate sites from the Site Table.
After simplification, since $\tsimplify{P', \Nt, \Et}$ results in the $N$ individuals in $P'$
being relabeled in \tskit{}'s tables as $0, 1, \ldots, N-1$,
there must be bookkeeping that keeps in sync the individual IDs as recorded by the simulator
with the node IDs recorded in the tables.

>>>>

\begin{abstract}
    In this paper we describe how to efficiently record the entire genetic history of a population
    in forwards-time, individual-based population genetics simulations with arbitrary breeding models,
    population structure and demography.
    This approach dramatically reduces the computational burden of tracking individual genomes
    by allowing us to simulate only those loci that may affect reproduction
    (those having non-neutral variants).
    The genetic history of the population is recorded as a succinct tree sequence
    as introduced in the software package \msprime,
    on which neutral mutations can be quickly placed afterwards.
    Recording the results of each breeding event requires storage that grows linearly with time,
    but there is a great deal of redundancy in this information.
    We solve this storage problem by providing an algorithm to quickly `simplify'
    a tree sequence by removing this irrelevant history for a given set of genomes.
    By periodically simplifying the history with respect to the extant population,
    we show that the total storage space required is modest
    and overall large efficiency gains can be made over classical forward-time simulations.
    We implement a general-purpose framework for recording and simplifying genealogical data,
    which can be used to make simulations of any population model more efficient.
    We modify two popular forwards-time simulation frameworks to use this new approach
    and observe efficiency gains in large, whole-genome simulations of one to two orders of magnitude.
    In addition to speed, our method for recording pedigrees has several advantages:
    (1) All marginal genealogies of the simulated individuals are recorded, rather than just genotypes.
    (2) A population of $N$ individuals with $M$ polymorphic sites
        can be stored in $O(N \log N + M)$ space,
        making it feasible to store a simulation's entire final generation
        as well as its history.
    (3) A simulation can easily be initialized with a more efficient coalescent simulation 
        of deep history.
    The software for recording and processing tree sequences is named \tskit.
\end{abstract}

@@

\paragraph{Essential steps in tree sequence recording}
Since a tree sequence can record the history of genetic inheritance in any situation
(requiring only unambiguous inheritance and no time travel),
any individual-based population genetics simulator can maintain a tree sequence
with only a little bookkeeping. \revpoint{1}{1}
We have furthermore provided several tools to minimize this bookkeeping,
only requiring one-way output at the birth of each new individual.
Concretely, to record a tree sequence, including mutations,
a simulator must record for each new genome
(so, twice for each new diploid individual): \revpoint{1}{6}
\begin{itemize}
    \item the birth time of the genome in the Node Table,
    \item the segments the genome inherits from its parental genomes in the Edge Table,
    \item the locations of any new mutations in the Site Table,
    \item and the derived state of these new mutations in the Mutation Table
        (as well as the identiy of this genome the mutations appeared in).
\end{itemize}
Each of these can be simply appended to the ends of the respective tables.
Besides this, simplification should be run every once in a while (e.g., every 100 generations).
Before simplification, time in the Node Table must be translated to ``time ago'' if it is not already.
(This was avoided in Algorithm~\algref{W} since there $t$ denoted
``time until the end of the simulation''.)
There are also several ``cleaning'' steps, for which we provide functions in the Tables API:
\emph{sorting} according to several criteria for algorithmic efficiency;
and removing any duplicate sites from the Site Table.
After simplification, since $\tsimplify{P', \Nt, \Et}$ results in the $N$ individuals in $P'$
being relabeled in \tskit{}'s tables as $0, 1, \ldots, N-1$,
there must be bookkeeping that keeps in sync the individual IDs as recorded by the simulator
with the node IDs recorded in the tables.
